---
title: "Llama Parse -> GPT-4"
author: "Marc Teunis"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


Step 1: Load the Necessary Package

```{r}

if (!requireNamespace("deSolve", quietly = TRUE)) {
    install.packages("deSolve")
}
library(deSolve)
```

## Step 2: Define the PBPK Model Function

This function will model the physiology-based pharmacokinetics using differential equations for each compartment (Liver, Fat, Brain, Lung, Kidney):

```{r}

pbpk_model <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
      # Equations for PFOA or PFOS concentration in each compartment
      dLiver = (Q_blood * free * (Ca - Liver/Kp_liver) + intake) / V_liver - k_met * Liver
      dFat = (Q_blood * free * (Ca - Fat/Kp_fat)) / V_fat
      dBrain = (Q_blood * free * (Ca - Brain/Kp_brain)) / V_brain
      dLung = (Q_blood * free * (Ca - Lung/Kp_lung)) / V_lung
      dKidney = (Q_blood * free * (Ca - Kidney/Kp_kidney) + Tm * Cfil/(Kt + Cfil)) / V_kidney
      
      # Return the rate of change
      list(c(dLiver, dFat, dBrain, dLung, dKidney))
    })
}
```

## Step 3: Parameters and Initial Conditions

This code sets the initial conditions and parameters for one of the scenarios, e.g., PFOA human-based. You would repeat and modify these for each chemical and species based on the partition coefficients and any other species-specific parameters.

```{r}

# PFOA Human
parameters <- c(Q_blood = 55.2,  # example blood flow rate (mL/h)
                free = 0.03,      # free fraction in plasma
                Ca = 0.5,         # arterial concentration example (pg/mL)
                Kp_liver = 1.03, 
                Kp_fat = 0.47, 
                Kp_brain = 0.17, 
                Kp_lung = 1.27, 
                Kp_kidney = 1.17,
                V_liver = 1.64, 
                V_fat = 1.43, 
                V_brain = 1.50,
                V_lung = 1.00, 
                V_kidney = 0.29,
                k_met = 0.1,      # metabolism rate constant
                intake = 0.004583333,    # intake rate (microgram/h)
                Tm = 147.4,       # resorption max (pg/h)
                Cfil = 0.03,      # concentration in filtrate (pg/mL)
                Kt = 0.116)       # affinity constant (ng/mL)

initial_conditions <- c(Liver = 0, Fat = 0, Brain = 0, Lung = 0, Kidney = 0)
```

Step 4: Solve the Differential Equations

```{r}

times <- seq(0, 175200, by = 8760)  # Simulate for 20 years, yearly intervals
results <- ode(y = initial_conditions, times = times, func = pbpk_model, parms = parameters)
```

## Step 5: Plot the Results Using ggplot2

```{r}

library(ggplot2)
results_df <- as.data.frame(results)
results_melted <- reshape2::melt(results_df, id.vars = "time")

ggplot(data = results_melted, aes(x = time, y = value, color = variable)) +
  geom_line() +
  labs(title = "PBPK Modeling of PFOA in Humans",
       x = "Time (hours)", y = "Concentration (microgram/mL)",
       color = "Compartment") + 
  facet_wrap(~variable)
```

This setup assumes you have predefined all the necessary physiological parameters. For the other three scenarios (PFOA rat-based, PFOS human-based, and PFOS rat-based), you'll adjust the partition coefficients and other physiological parameters as necessary.

Each model will help you understand how PFOS and PFOA distribute across different body compartments under different physiological conditions, potentially supporting risk assessment and regulatory decisions.

```{r}
# Load necessary libraries
if (!requireNamespace("deSolve", quietly = TRUE)) {
    install.packages("deSolve")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
    install.packages("ggplot2")
}
library(deSolve)
library(ggplot2)

# Define the PBPK model function
pbpk_model <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
      # Equations for concentration in each compartment
      dLiver = (Q_blood * free * (Ca - Liver/Kp_liver) + intake) / V_liver - k_met * Liver
      dFat = (Q_blood * free * (Ca - Fat/Kp_fat)) / V_fat
      dBrain = (Q_blood * free * (Ca - Brain/Kp_brain)) / V_brain
      dLung = (Q_blood * free * (Ca - Lung/Kp_lung)) / V_lung
      dKidney = (Q_blood * free * (Ca - Kidney/Kp_kidney) + Tm * Cfil/(Kt + Cfil)) / V_kidney
      
      # Return the rate of change
      list(c(dLiver, dFat, dBrain, dLung, dKidney))
    })
}

# Parameters and initial conditions for all four scenarios
params_list <- list(
  rat_PFOA = c(Q_blood = 55.2, free = 0.03, Ca = 0.5, intake = 0.13, 
               Kp_liver = 2.20, Kp_fat = 0.04, Kp_brain = 0.01, Kp_lung = 0.15, Kp_kidney = 1.05,
               V_liver = 1.64, V_fat = 1.43, V_brain = 1.50, V_lung = 1.00, V_kidney = 0.29,
               k_met = 0.1, Tm = 147.4, Cfil = 0.03, Kt = 0.116),
  human_PFOA = c(Q_blood = 55.2, free = 0.03, Ca = 0.5, intake =  0.004583333,
                 Kp_liver = 1.03, Kp_fat = 0.47, Kp_brain = 0.17, Kp_lung = 1.27, Kp_kidney = 1.17,
                 V_liver = 1.64, V_fat = 1.43, V_brain = 1.50, V_lung = 1.00, V_kidney = 0.29,
                 k_met = 0.1, Tm = 147.4, Cfil = 0.03, Kt = 0.116),
  rat_PFOS = c(Q_blood = 55.2, free = 0.03, Ca = 0.5, intake = 0.13,
               Kp_liver = 3.72, Kp_fat = 0.14, Kp_brain = 0.01, Kp_lung = 0.15, Kp_kidney = 0.80,
               V_liver = 1.64, V_fat = 1.43, V_brain = 1.50, V_lung = 1.00, V_kidney = 0.29,
               k_met = 0.1, Tm = 147.4, Cfil = 0.03, Kt = 0.116),
  human_PFOS = c(Q_blood = 55.2, free = 0.03, Ca = 0.5, intake = 0.005416667,
                 Kp_liver = 2.67, Kp_fat = 0.33, Kp_brain = 0.26, Kp_lung = 0.15, Kp_kidney = 1.26,
                 V_liver = 1.64, V_fat = 1.43, V_brain = 1.50, V_lung = 1.00, V_kidney = 0.29,
                 k_met = 0.1, Tm = 147.4, Cfil = 0.03, Kt = 0.116)
)

initial_conditions <- c(Liver = 0, Fat = 0, Brain = 0, Lung = 0, Kidney = 0)
times <- seq(0, 175200, by = 8760)  # Simulation for 3 days at hourly intervals

# Simulate all scenarios
results_list <- lapply(names(params_list), function(scenario) {
  ode(y = initial_conditions, times = times, func = pbpk_model, parms = params_list[[scenario]])
})

## function to plot
names(results_list) <- c("rat-pfoa", "human-pfoa", "rat-pfos", "human-pfos")
df = results_list[[1]]
plot_pk <-  function(df, species){
  
  df <- df |> 
    as_tibble() |>
    janitor::clean_names() |>
    gather()
    
  
}


# Prepare data for plotting
results_df <- do.call(rbind, lapply(1:length(results_list), function(i) {
  as.data.frame(results_list[[i]]) %>%
  reshape2::melt(id.vars = "time") %>% 
  dplyr::mutate(Scenario = names(results_list)[i])
}))

## convert to years
results_df <- results_df |>
  mutate(years = time/8760)

## nest 
df_nested <- results_df |>
  group_by(Scenario, variable) |>
  nest()




# Plotting using ggplot2
plot_this <- function(df, title){
 p <- ggplot(data = df, aes(x = years, y = value)) +
  geom_line() +
 # facet_grid(Scenario ~ variable) +
  labs(title = title,
       x = "Time (years)", y = "Concentration (ng/g)") +
  theme_minimal()
p
  
}



## apply plot to nested
df_nested <- df_nested |>
  dplyr::mutate(
    plots = map2(
      data,
      paste(Scenario, variable),
      plot_this))

df_nested$plots
```

